### 2020-12-31 written/eLifeRevision/Bittner_eLife2020.tex
SC5 caption
tex updates

\thispagestyle{plain}
{\Large Interrogating theoretical models of neural computation with deep inference} \\
Sean R. Bittner$^{1}$, Agostina Palmigiano$^{1}$, Alex T. Piet$^{2,3}$, Chunyu A. Duan$^{4}$, Carlos D. Brody$^{2,3,5}$, \\
Kenneth D. Miller$^{1}$, and John P. Cunningham$^{6}$.

{\small
$^{1}$Department of Neuroscience, Columbia University, \\
$^{2}$Princeton Neuroscience Institute, \\
$^{3}$Princeton University, \\
$^{4}$Institute of Neuroscience, Chinese Academy of Sciences, \\
$^{5}$Howard Hughes Medical Institute, \\
$^{6}$Department of Statistics, Columbia University
}

\linenumbers
\section{Abstract}
A cornerstone of theoretical neuroscience is the circuit model: a system of equations that captures a hypothesized neural mechanism.  
Such models are valuable when they give rise to an experimentally observed phenomenon -- whether behavioral or in terms of neural activity -- and thus can offer insights into neural computation.
The operation of these circuits, like all models, critically depends on the choices of model parameters.
When analytic derivation of the relationship between model parameters and computational properties is intractable, approximate inference and simulation-based techniques are relied upon for scientific insight.
%Historically, the gold standard has been to analytically derive the relationship between model parameters and computational properties.  
%However, this enterprise quickly becomes infeasible as biologically realistic constraints are included into the model increasing its complexity, often resulting in \emph{ad hoc} approaches to understanding the relationship between model and computation.
We bring the use of deep generative models for probabilistic inference to bear on this problem, learning distributions of parameters that produce the specified properties of computation.
By learning parameter distributions that produce computations -- an emergent property, we introduce a novel methodology for exploratory analyses and hypothesis testing that is particularly well-suited to the stochastic dynamical systems models predominant in our field of theoretical neuroscience.
%Importantly, the techniques we introduce offer a principled means to understand the implications of model parameter choices on computational properties of interest.  
We motivate this methodology with a worked example analyzing sensitivity in the stomatogastric ganglion.  
We then use it to go beyond linear theory of neuron-type input-responsivity in a model of primary visual cortex, gain a mechanistic understanding of rapid task switching in superior colliculus models, and attribute error to connectivity properties in recurrent neural networks solving a simple mathematical task.
While much use of deep learning in theoretical neuroscience focuses on drawing analogies between optimized neural architectures and the brain, this work illustrates how we can further leverage the power of deep learning towards solving inverse problems in theoretical neuroscience.
%More generally, this work suggests a departure from realism vs tractability considerations, towards the use of modern machine learning for sophisticated interrogation of biologically relevant models.

@@

% What's wrong with existing methodology:
Existing approaches to the inverse problem in theoretical neuroscience fall short in three key ways. 
First, theoretical models of neural computation aim to reflect a complex biological reality, and as a result, such models lack tractable likelihoods.
Thus, standard approaches from statistical inference are unavailable.
The parameter sets obtained from likelihood-free ABC lack a formalized link to Bayesian inference (except in the unrealistic $0$-distance scenario), lack parameter probabilities, and only confer sensitivity analyses of an alternative likelihood to the simulator-defined likelihood of ABC \cite{raue2009structural}.
Second is the undesirable trade-off between the flexibility and tractability of the approximated posterior distribution.
While sampling-based approaches like ABC and Markov chain Monte Carlo (MCMC) can produce flexible posterior approximations, they must be run continually for increasing samples.
While VAE approaches can result in tractable posterior sampling and sensitivity measurements post-optimization, existing approaches have relied on simplified classes of distributions, which restrict the flexibility of the posterior approximation.
And third, you can never make assumptions of what inferred model parameters will predict.
This is well understood when considering Box's loop and the role of posterior predictive checks in the development and critique of scientific models \cite{gelman2013philosophy, blei2014build}.
Uncertainty about the properties of inferred model predictions introduce a conceptual degree of freedom to the inverse problem that may be unnecessary and undesirable given the scientific motivation.

@@

Equipped with this methodology, we then investigated three models of current importance in theoretical neuroscience.
These models were chosen to demonstrate generality through ranges of biological realism (from conductance-based biophysics to recurrent neural networks), neural system function (from pattern generation to abstract cognitive function), and network scale (from four to infinite neurons).
%First, we use EPI to produce a set of verifiable hypotheses of input-responsivity in a four neuron-type dynamical model of primary visual cortex; we then validate these hypotheses in the model.
First, we use EPI to elucidate the mechanisms of inhibition stabilization with varying contrast in a stochastic nonlinear dynamical model of primary visual cortex with inhibitory multiplicity.
%Second, we demonstrated how the systematic application of EPI to levels of task performance can generate experimentally testable hypotheses regarding connectivity in superior colliculus.  
Second, we discover connectivity patterns in superior colliculus resulting in resilience to optogenetic perturbation by using EPI to condition on rapid task switching.
Third, we use EPI to uncover the sources of error in a low-rank recurrent neural network executing a simple mathematical task.  
The novel scientific insights offered by EPI contextualize and clarify the previous studies exploring these models \cite{gutierrez2013multiple, litwin2016inhibitory, duan2018collicular, mastrogiuseppe2018linking}, and more generally, these results point to the value of deep inference for the interrogation of biologically relevant models.

@@

Parameters are represented by the vector $\mathbf{z}$ and data by the vector $\mathbf{x}$.
\textbf{B}. Simulated activity form the STG model at $g_{\text{el}} = 4.5$nS and $g_{\text{synA}} = 3$nS.
\textbf{C}. The emergent property of unified intermediacy, in which all neurons are firing close to the same intermediate frequency.
Simulated activity traces are colored by log probability density of their generating parameters in the EPI-inferred distribution. 
\textbf{D}. For a choice of model and emergent property, emergent property inference (EPI) learns a distribution of the model parameters $\mathbf{z} = \left[g_{\text{el}}, g_{\text{synA}} \right]$ producing middle hub frequency.

@@

\textbf{E}. The EPI distribution of STG model parameters producing network syncing.  

@@

$v_1$ is sensitive to network syncing ($p<10^{-4}$), while $v_2$ is not ($p=0.67$) (see Section \ref{methods_STG}). 

@@

\label{fig:STG}

@@

%Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}A gray box) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution shown in Figure \ref{fig:STG}B.
Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}D) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution  EPI (Fig. \ref{fig:STG}E).  
This distribution -- represented for clarity as samples from the distribution -- is then a scientifically meaningful and mathematically tractable object.  
In the STG model, this distribution can be specifically queried to reveal the prototypical parameter configuration for network syncing (the mode; Figure \ref{fig:STG}E yellow star), and how network syncing decays based on changes away from the mode.
The eigenvectors (of the Hessian of the distribution at the mode) quantitatively formalize the robustness of unified intermediacy (Fig. \ref{fig:STG}B solid ($v_1$) and dashed  ($v_2$) black arrows).
Indeed, samples equidistant from the mode along these EPI-identified dimensions of sensitivity ($v_1$) and degeneracy ($v_2$) agree with error contours (Fig. \ref{fig:STG}B contours) and have diminished or preserved network syncing, respectively (Fig. \ref{fig:STG}F activity traces, Fig. S TODO) (see Section \ref{methods_STG}).

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{figures/figSC5/figSC5.pdf}
\end{center}
\caption{\small (SC5): Caption...
}
\label{fig:SC5}
\end{figure}

>>>>

\thispagestyle{plain}
{\Large Interrogating theoretical models of neural computation with deep inference} \\
Sean R. Bittner$^{1}$, Agostina Palmigiano$^{1}$, Alex T. Piet$^{2,3,4}$, Chunyu A. Duan$^{5}$, Carlos D. Brody$^{2,3,6}$, \\
Kenneth D. Miller$^{1}$, and John P. Cunningham$^{7}$.

{\small
$^{1}$Department of Neuroscience, Columbia University, \\
$^{2}$Princeton Neuroscience Institute, \\
$^{3}$Princeton University, \\
$^{4}$Allen Institute for Brain Science, \\
$^{5}$Institute of Neuroscience, Chinese Academy of Sciences, \\
$^{6}$Howard Hughes Medical Institute, \\
$^{7}$Department of Statistics, Columbia University
}

\linenumbers
\section{Abstract}
A cornerstone of theoretical neuroscience is the circuit model: a system of equations that captures a hypothesized neural mechanism.  
Such models are valuable when they give rise to an experimentally observed phenomenon -- whether behavioral or in terms of neural activity -- and thus can offer insights into neural computation.
The operation of these circuits, like all models, critically depends on the choices of model parameters.
When analytic derivation of the relationship between model parameters and computational properties is intractable, approximate inference and simulation-based techniques are relied upon for scientific insight.
%Historically, the gold standard has been to analytically derive the relationship between model parameters and computational properties.  
%However, this enterprise quickly becomes infeasible as biologically realistic constraints are included into the model increasing its complexity, often resulting in \emph{ad hoc} approaches to understanding the relationship between model and computation.
We bring the use of deep generative models for probabilistic inference to bear on this problem, learning distributions of parameters that produce the specified properties of computation.
By learning parameter distributions that produce computations -- an emergent property, we introduce a novel methodology that is particularly well-suited to the stochastic dynamical systems models predominant in our field of theoretical neuroscience.
%Importantly, the techniques we introduce offer a principled means to understand the implications of model parameter choices on computational properties of interest.  
We motivate this methodology with a worked example analyzing sensitivity in the stomatogastric ganglion.  
%We then use it to go beyond linear theory of neuron-type input-responsivity in a model of primary visual cortex, gain a mechanistic understanding of rapid task switching in superior colliculus models, and attribute error to connectivity properties in recurrent neural networks solving a simple mathematical task.
We then use it to reveal the key factors of variability in a model of primary visual cortex, gain a mechanistic understanding of rapid task switching in superior colliculus models, and scale inference of large low-rank RNN's exhibiting stable amplification.
While much use of deep learning in theoretical neuroscience focuses on drawing analogies between optimized neural architectures and the brain, this work illustrates how we can further leverage the power of deep learning towards solving inverse problems in theoretical neuroscience.
%More generally, this work suggests a departure from realism vs tractability considerations, towards the use of modern machine learning for sophisticated interrogation of biologically relevant models.

@@

% What's wrong with existing methodology:
Existing approaches to the inverse problem in theoretical neuroscience fall short in three key ways. 
First, theoretical models of neural computation aim to reflect a complex biological reality, and as a result, such models lack tractable likelihoods.
Thus, standard approaches from statistical inference are unavailable.
The parameter sets obtained from likelihood-free ABC lack a formalized link to Bayesian inference (except in the unrealistic $0$-distance scenario), and lack parameter probabilities.
Second is the undesirable trade-off between the flexibility and tractability of the approximated posterior distribution.
While sampling-based approaches like ABC and Markov chain Monte Carlo (MCMC) can produce flexible posterior approximations, they must be run continually for increasing samples.
While VAE approaches can result in tractable posterior sampling and sensitivity measurements post-optimization, existing approaches have relied on simplified classes of distributions, which restrict the flexibility of the posterior approximation.
And third, you can never make assumptions of what inferred model parameters will predict.
This is well understood when considering Box's loop and the role of posterior predictive checks in the development and critique of scientific models \cite{gelman2013philosophy, blei2014build}.
Uncertainty about the properties of inferred model predictions introduce a conceptual degree of freedom to the inverse problem that may be unnecessary and undesirable given the scientific motivation.

@@

Equipped with this methodology, we then investigated three models of current importance in theoretical neuroscience.
These models were chosen to demonstrate generality through ranges of biological realism (from conductance-based biophysics to recurrent neural networks), neural system function (from pattern generation to abstract cognitive function), and network scale (from four to 100 neurons).
%First, we use EPI to produce a set of verifiable hypotheses of input-responsivity in a four neuron-type dynamical model of primary visual cortex; we then validate these hypotheses in the model.
First, we use EPI to understand the characteristics of noise that govern Fano factor in a stochastic four neuron-type model of primary visual cortex.
%Second, we demonstrated how the systematic application of EPI to levels of task performance can generate experimentally testable hypotheses regarding connectivity in superior colliculus.  
Second, we discover connectivity patterns in superior colliculus resilient to optogenetic perturbation by using EPI to condition on rapid task switching.
The novel scientific insights offered by EPI contextualize and clarify the previous studies exploring these models \cite{litwin2016inhibitory, duan2018collicular}.
%Third, we use EPI to uncover the sources of error in a low-rank recurrent neural network executing a simple mathematical task.  
Third, we emphasize the methodological advancement of EPI by inferring high-dimensional distributions of RNN connectivities exhibiting stable amplification.
These results point to the value of deep inference for the interrogation of biologically relevant models.

@@

Parameters are represented by the vector $\mathbf{z}$ and membrane potentials by the vector $\mathbf{x}$.
The evolution of this model's activity $\mathbf{x}(t)$ is predicated by differential equations.
\textbf{B}. Spiking frequency $\omega(\mathbf{x}; \mathbf{z})$ is an emergent property statistic. Spiking frequency is measured from simulated activity of the STG model at parameter choices of $g_{\text{el}} = 4.5$nS and $g_{\text{synA}} = 3$nS.
\textbf{C}. The emergent property of intermediate hub frequency, in which the hb neuron fires at a rate between the fast and slow frequencies.
Simulated activity traces are colored by log probability density of their generating parameters in the EPI-inferred distribution (Panel E).
\textbf{D}. For a choice of model and emergent property, emergent property inference (EPI) learns a distribution of the model parameters $\mathbf{z} = \left[g_{\text{el}}, g_{\text{synA}} \right]$ producing intermediate hub frequency.

@@

\textbf{E}. The EPI distribution of STG model parameters producing intermediate hub frequency.

@@

@@

\label{fig:STG_EPI}

@@

%Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}A gray box) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution shown in Figure \ref{fig:STG}B.
Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG_EPI}D) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution  EPI (Fig. \ref{fig:STG}E).  
This distribution -- represented for clarity as samples from the distribution -- is then a scientifically meaningful and mathematically tractable object.  
In the STG model, this distribution can be specifically queried to reveal the prototypical parameter configuration for network syncing (the mode; Figure \ref{fig:STG}E yellow star), and how network syncing decays based on changes away from the mode.
The eigenvectors (of the Hessian of the distribution at the mode) quantitatively formalize the robustness of unified intermediacy (Fig. \ref{fig:STG}B solid ($v_1$) and dashed  ($v_2$) black arrows).
Indeed, samples equidistant from the mode along these EPI-identified dimensions of sensitivity ($v_1$) and degeneracy ($v_2$) agree with error contours (Fig. \ref{fig:STG}B contours) and have diminished or preserved network syncing, respectively (Fig. \ref{fig:STG}F activity traces, Fig. S TODO) (see Section \ref{methods_STG}).

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.6]{figures/figSC5/figSC5.pdf}
\end{center}
\caption{\small (SC5): EPI optimization of the SC model producing rapid task switching.
A. Entropy throughout optimization. 
B. The emergent property statistic means and variances converge to their constraints at 12,000 iterations following the sixth augmented Lagrangian epoch.
}
\label{fig:SC5}
\end{figure}
